Graph embedding and transfer learning can help predict potential species interaction networks despite data limitations

### 2023-09-07 docs/_tex/index.tex
ready

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Graph embedding and transfer learning can help predict potential species interaction networks despite data limitations},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Graph embedding and transfer learning can help predict potential
species interaction networks despite data limitations}
\author{Tanya Strydom \and Salomé Bouskila \and Francis
Banville \and Ceres Barros \and Dominique Caron \and Maxwell J.
Farrell \and Marie-Josée Fortin \and Victoria Hemming \and Benjamin
Mercier \and Laura J. Pollock \and Rogini Runghen \and Giulio V. Dalla
Riva \and Timothée Poisot}
\date{}

\begin{document}
\maketitle
The ability to infer \emph{potential} interactions could serve as a
significant breakthrough in our ability to conceptualize species
interaction networks over large spatial scales (Hortal et al. 2015).
Reliable inferences would not only boost our understanding of the
structure of species interaction networks, but also increase the amount
of information that can be used for biodiversity management. In a recent
overview of the field of ecological network prediction, Strydom et al.
(2021) identified two challenges of interest to the prediction of
interactions at large scales. First, there is a relative scarcity of
relevant data in most places globally -- which, due to the limitations
in most predictive methods, restricts the ability to infer interactions
to locations where it is least required (\emph{i.e.} regions where we
already have interaction data) leaving us unable to make inference in
data scarce regions (where we most need it); second, accurate predictors
are important for accurate predictions, and the lack of methods that can
leverage a small amount of \emph{accurate} data is a serious impediment
to our predictive ability. In this contribution, we highlight the power
of viewing (and constructing) metawebs as \emph{probabilistic} objects
in the context of low-probability interactions, discuss how a family of
machine learning tools (graph embeddings and transfer learning) can be
used to overcome data limitations to metaweb inference, and highlight
how the use of metawebs introduces important questions for the field of
network ecology.

In most places, our most reliable biodiversity knowledge is that of a
species pool where a set of potentially interacting species in a given
area could occur: through the analysis of databases like the Global
Biodiversity Information Facility (GBIF) or the International Union for
the Conservation of Nature (IUCN), it is possible to construct a list of
species for a region of interest. Following the definition of Dunne
(2006), a metaweb is the ecological network analogue to the species
pool; specifically, it inventories all \emph{potential} interactions
between species for a spatially delimited area (and so captures the
\(\gamma\) diversity of interactions as per Poisot et al. (2012)).
However, inferring the potential interactions between these species
still remains a challenge. And yet, because the metaweb represents the
joint effect of functional, phylogenetic, and macroecological processes
(Morales-Castilla et al. 2015; Carlson et al. 2022; Morales-Castilla et
al. 2021), it holds valuable ecological information. Specifically, it
represents the ``upper bounds'' on what the composition of the local
networks, given a local species pool, can be (see \emph{e.g.} McLeod et
al. 2021); this information can help evaluate the ability of ecological
assemblages to withstand the effects of, for example, climate change
(Fricke et al. 2022). These local networks may be reconstructed given an
appropriate knowledge of local species composition and provide
information on the structure of networks at finer spatial scales. This
has been done for example for tree-galler-parasitoid systems (Gravel et
al. 2018), fish trophic interactions (Albouy et al. 2019), tetrapod
trophic interactions (J. Braga et al. 2019; O'Connor et al. 2020), and
crop-pest networks (Grünig et al. 2020).

The metaweb itself is not a prediction of local networks at specific
locations within the spatial area it covers: it will have a different
structure, notably by having a larger connectance (see \emph{e.g.} Wood
et al. 2015) and complexity (see \emph{e.g.} Galiana et al. 2022), than
any of these local networks. These local networks (which capture the
\(\alpha\) diversity of interactions) are a subset of the metaweb's
species and its realized interactions, and have been called ``metaweb
realizations'' (Poisot, Stouffer, and Gravel 2015). Differences between
local networks and their metawebs are due to chance, species abundance
and co-occurrence, local environmental conditions, and local
distribution of functional traits, among others. Specifically, although
co-occurrence can be driven by interactions (Cazelles et al. 2016),
co-occurrence alone is not a predictor of interactions (Blanchet,
Cazelles, and Gravel 2020; Thurman et al. 2019), and therefore the lack
of co-occurrence cannot be used to infer the lack of a feasible
interaction. Yet, recent results by Saravia et al. (2021) strongly
suggested that local (metaweb) realizations only respond weakly to local
conditions: instead, they reflect constraints inherited by the structure
of their metaweb. This sets up the core goal of predictive network
ecology as the prediction of metaweb structure, as it is required to
accurately produce downscaled, local predictions.

@@

Treating interactions as probabilistic (as opposed to binary) events is
a more nuanced and realistic way to represent them. Dallas, Park, and
Drake (2017) suggested that most interactions (links) in ecological
networks are cryptic, \emph{i.e.} uncommon or hard to observe. This
argument echoes Jordano (2016): sampling ecological interactions is
difficult because it requires first the joint observation of two
species, and then the observation of their interaction. In addition, it
is generally expected that weak or rare interactions will be more
prevalent in networks than common or strong interactions (Csermely
2004), compared to strong, persistent interactions; this is notably the
case in food chains, wherein many weaker interactions are key to the
stability of a system (Neutel, Heesterbeek, and de Ruiter 2002). In the
light of these observations, we expect to see an over-representation of
low-probability (hereafter rare) interactions under a model that
accurately predicts interaction probabilities.

Yet, the original metaweb definition, and indeed most past uses of
metawebs, was based on the presence/absence of interactions. Moving
towards \emph{probabilistic} metawebs, by representing interactions as
Bernoulli events (see \emph{e.g.} Poisot et al. 2016), offers the
opportunity to weigh these rare interactions appropriately. The inherent
plasticity of interactions is important to capture: there have been
documented instances of food webs undergoing rapid collapse/recovery
cycles over short periods of time (\emph{e.g.} Pedersen et al. 2017).
Furthermore, because the structure of the metaweb cannot be known in
advance, it is important to rely on predictive tools that do not assume
a specific network topology for link prediction (Gaucher, Klopp, and
Robin 2021), but are able to work on generalizations of the network.
These considerations emphasize why metaweb predictions should focus on
quantitative (preferentially probabilistic) predictions, and this should
constrain the suite of models that are appropriate for prediction.
Binary classifiers based on probabilities have an extremely robust
methodology to validate them, and this applies naturally to the
prediction of interactions (Poisot 2023).

It is important to recall that a metaweb is intended as a catalogue of
all potential (feasible) interactions, which is then filtered for a
given application (Morales-Castilla et al. 2015). It is therefore
important to separate the interactions that happen ``almost surely''
(repeated observational data), ``almost never'' (repeated lack of
evidence \emph{or} evidence that the link is forbidden through
\emph{e.g.} trait mis-match), and interactions with a probability that
lays somewhere in between (Catchen et al. 2023). Although metawebs can
(and in practice likely do) include false positives, these are
statistically negligible compared to the false negatives. Furthermore,
Strydom et al. (2022) shows that t-SVD embedding is extremely robust to
(and able to detect) the presence of false positives. In a sense, that
most ecological interactions are elusive can call for a slightly
different approach to sampling: once the common interactions are
documented, the effort required in documenting each rare interaction
will increase exponentially (Jordano 2016). Recent proposals in other
fields relying on machine learning approaches emphasize the idea that
algorithms meant to predict, through the assumption that they
approximate the process generating the data, can also act as data
generators (Hoffmann et al. 2019). High quality observational data can
be used to infer core rules underpinning network structure, and be
supplemented with synthetic data coming from predictive models trained
on them, thereby increasing the volume of information available for
analysis. Indeed, Strydom et al. (2021) suggested that knowing the
metaweb may render the prediction of local networks easier, because it
fixes an ``upper bound'' on which interactions can exist. In this
context, a probabilistic metaweb represents an aggregation of
informative priors on the biological feasibility of interactions, which
is usually hard to obtain yet has possibly the most potential to boost
our predictive ability of local networks (Bartomeus 2013; Bartomeus et
al. 2016). This would represent a departure from simple rules expressed
at the network scale (\emph{e.g.} Williams and Martinez 2000) to a view
of network prediction based on learning the rules that underpin
interactions \emph{and} their variability (Anubhav Gupta, Furrer, and
Petchey 2022).

@@

Graph (or network) embedding (Figure~\ref{fig-embedding}) is a family of
machine learning techniques, whose main task is to learn a mapping
function from a discrete graph to a continuous domain (Arsov and Mirceva
2019; Chami et al. 2022). Their main goal is to learn a low dimensional
vector representation of the graph (embeddings), such that its key
properties (\emph{e.g.} local or global structures) are retained in the
embedding space (Yan et al. 2005). The embedding space may, but will not
necessarily, have lower dimensionality than the graph. Ecological
networks are promising candidates for the routine application of
embeddings, as they tend to possess a shared structural backbone (see
\emph{e.g.} Bramon Mora et al. 2018), which hints at structural
invariants in empirical data. Assuming that these structural invariants
are common enough, they would dominate the structure of networks, and
therefore be adequately captured by the first (lower) dimensions of an
embedding, without the need to measure derived aspects of their
structure (\emph{e.g.} motifs, paths, modularity, \ldots).

@@

Rather than directly predicting biological rules (see \emph{e.g.}
Pichler et al. 2020 for an overview), which may be confounded by the
sparse nature of graph data, learning embeddings works in the
low-dimensional space that maximizes information about the network
structure. This approach is further justified by the observation, for
example, that the macro-evolutionary history of a network is adequately
represented by some graph embeddings {[}Random dot product graphs
(RDPG); see Dalla Riva and Stouffer (2016){]}. In a recent publication,
Strydom et al. (2022) have used an embedding (based on RDPG) to project
a metaweb of trophic interactions between European mammals, and
transferred this information to mammals of Canada, using the
phylogenetic distance between related clades to infer the values in the
latent subspace into which the European metaweb was projected. By
performing the RDPG step on re-constructed values, this approach yields
a probabilistic trophic metaweb for mammals of Canada based on knowledge
of European species, despite a limited (\(\approx\) 5\%) taxonomic
overlap, and illustrates how the values derived from an embedding can be
used for prediction without being ``traits'' of the species they
represent.

@@

Graph embeddings \emph{can} serve as a dimensionality reduction method.
For example, RDPG (Strydom et al. 2022) and t-SVD {[}truncated Singular
Value Decomposition; Poisot et al. (2021){]} typically embed networks
using fewer dimensions than the original network {[}the original network
has as many dimensions as species, and as many informative dimensions as
trophically unique species; Strydom, Dalla Riva, and Poisot (2021){]}.
However, this is not necessarily the case -- indeed, one may perform a
PCA (a special case of SVD) to project the raw data into a subspace that
improves the efficacy of t-SNE {[}t-distributed stochastic neighbor
embedding; Maaten (2009){]}. There are many dimensionality reductions
(Anowar, Sadaoui, and Selim 2021) that can be applied to an embedded
network should the need for dimensionality reduction (for example for
data visualization) arise. In brief, many graph embeddings \emph{can}
serve as dimensionality reduction steps, but not all do, neither do all
dimensionality reduction methods provide adequate graph embedding
capacities. In the next section (and Figure~\ref{fig-embedding}), we
show how the amount of dimensionality reduction can affect the quality
of the embedding.

\hypertarget{tbl-methods}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{IBM Plex Serif}
  \setsansfont[]{IBM Plex Sans}
  \setmonofont[]{IBM Plex Mono}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Graph embedding and transfer learning can help predict potential species interaction networks despite data limitations},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Graph embedding and transfer learning can help predict potential
species interaction networks despite data limitations}
\author{Tanya Strydom %
%
\textsuperscript{%
1,%
2%
}%
; Salomé Bouskila %
%
\textsuperscript{%
%
1%
}%
; Francis Banville %
%
\textsuperscript{%
1,2,%
3%
}%
; Ceres Barros %
%
\textsuperscript{%
%
4%
}%
; Dominique Caron %
%
\textsuperscript{%
5,%
2%
}%
; Maxwell J. Farrell %
%
\textsuperscript{%
%
6%
}%
; Marie-Josée Fortin %
%
\textsuperscript{%
%
6%
}%
; Benjamin Mercier %
%
\textsuperscript{%
3,%
2%
}%
; Laura J. Pollock %
%
\textsuperscript{%
5,%
2%
}%
; Rogini Runghen %
%
\textsuperscript{%
%
7%
}%
; Giulio V. Dalla Riva %
%
\textsuperscript{%
%
7%
}%
; Timothée Poisot %
%
\textsuperscript{%
1,%
2%
}%
}
\date{}

\usepackage{setspace}
\usepackage[left,pagewise]{lineno}
\usepackage[letterpaper]{geometry}

\usepackage[nolists,noheads,markers]{endfloat}
\geometry{margin=2.5cm}

\begin{document}

\thispagestyle{empty}
{\bfseries\sffamily\Large Graph embedding and transfer learning can help
predict potential species interaction networks despite data limitations}
\vfil
Tanya Strydom %
%
\textsuperscript{%
1,%
2%
}%
; Salomé Bouskila %
%
\textsuperscript{%
%
1%
}%
; Francis Banville %
%
\textsuperscript{%
1,2,%
3%
}%
; Ceres Barros %
%
\textsuperscript{%
%
4%
}%
; Dominique Caron %
%
\textsuperscript{%
5,%
2%
}%
; Maxwell J. Farrell %
%
\textsuperscript{%
%
6%
}%
; Marie-Josée Fortin %
%
\textsuperscript{%
%
6%
}%
; Benjamin Mercier %
%
\textsuperscript{%
3,%
2%
}%
; Laura J. Pollock %
%
\textsuperscript{%
5,%
2%
}%
; Rogini Runghen %
%
\textsuperscript{%
%
7%
}%
; Giulio V. Dalla Riva %
%
\textsuperscript{%
%
7%
}%
; Timothée Poisot %
%
\textsuperscript{%
1,%
2%
}%

\vfil
{\small
\textbf{Abstract:} \begin{itemize}
\tightlist
\item
  Metawebs (networks of potential interactions within a species pool)
  are a powerful abstraction to understand how large-scale species
  interaction networks are structured.
\item
  Because metawebs are typically expressed at large spatial and
  taxonomic scales, assembling them is a tedious and costly process;
  predictive methods can help circumvent the limitations in data
  deficiencies, by providing a first approximation of metawebs.''
\item
  One way to improve our ability to predict metawebs is to maximize
  available information by using graph embeddings, as opposed to an
  exhaustive list of species interactions. Graph embedding is an
  emerging field in machine learning that holds great potential for
  ecological problems.
\item
  Here, we outline how the challenges associated with inferring metawebs
  line-up with the advantages of graph embeddings; followed by a
  discussion as to how the choice of the species pool has consequences
  on the reconstructed network, specifically as to the role of
  human-made (or arbitrarily assigned) boundaries and how these may
  influence ecological hypotheses.
\end{itemize}
\vfil
}
\clearpage
\setcounter{page}{1}
\doublespacing
\linenumbers
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The ability to infer potential biotic interactions could serve as a
significant breakthrough in our ability to conceptualize networks over
large spatial scales (Hortal et al., 2015). Reliable inferences would
not only boost our understanding of the structure of species interaction
networks, but also increase the amount of information that can be used
for biodiversity management. In a recent overview of the field of
ecological network prediction, Strydom, Catchen, et al. (2021)
identified two challenges of interest to the prediction of interactions
at large scales. First, there is a relative scarcity of relevant data in
most places globally -- which, due to the limitations in most predictive
methods, restricts the ability to infer interactions to locations where
it is least required (\emph{i.e.} regions where we already have
interaction data) leaving us unable to make inference in data scarce
regions (where we most need it); second, accurate predictors are
important for accurate predictions, and the lack of methods that can
leverage a small amount of \emph{accurate} data is a serious impediment
to our predictive ability. In this contribution, we (i) highlight the
power of viewing (and constructing) metawebs as \emph{probabilistic}
objects in the context of low-probability interactions, (ii) discuss how
a family of machine learning tools (graph embeddings and transfer
learning) can be used to overcome data limitations to metaweb inference,
and (iii) highlight how the use of metawebs introduces important
questions for the field of network ecology.

In most places, our most reliable biodiversity knowledge is that of a
species pool where a set of potentially interacting species in a given
area could occur: through the analysis of databases like the Global
Biodiversity Information Facility (GBIF) or the International Union for
the Conservation of Nature (IUCN), it is possible to construct a list of
species for a region of interest. Following the definition of Dunne
(2006), a metaweb is the ecological network analogue to the species
pool; specifically, it inventories all \emph{potential} interactions
between species for a spatially delimited area (and so captures the
\(\gamma\) diversity of interactions as per Poisot et al. (2012)).
However, inferring the potential interactions between these species
still remains a challenge. And yet, the metaweb holds valuable
ecological information: it represents the joint effect of functional,
phylogenetic, and macroecological processes (Carlson et al., 2022;
Morales-Castilla et al., 2015; Morales-Castilla et al., 2021).
Specifically, it represents the ``upper bounds'' on what the composition
of the local networks, given a local species pool, can be (see
\emph{e.g.} McLeod et al., 2021); this information can help evaluate the
ability of ecological assemblages to withstand the effects of, for
example, climate change (Fricke et al., 2022). These local networks may
be reconstructed given an appropriate knowledge of local species
composition and provide information on the structure of networks at
finer spatial scales. This has been done for example for
tree-galler-parasitoid systems (Gravel et al., 2018), fish trophic
interactions (Albouy et al., 2019), terrestrial tetrapod trophic
interactions (J. Braga et al., 2019; O'Connor et al., 2020), and
crop-pest networks (Grünig et al., 2020).

The metaweb itself is not a prediction of local networks at specific
locations within the spatial area it covers: it will have a different
structure, notably by having a larger connectance (see \emph{e.g.} Wood
et al., 2015) and complexity (see \emph{e.g.} Galiana et al., 2022),
than any of these local networks. Local networks (which capture the
\(\alpha\) diversity of interactions) are a subset of the metaweb's
species and its realized interactions, and have been called ``metaweb
realizations'' (Poisot et al., 2015). Differences between local networks
and their metawebs are due to chance, species abundance and
co-occurrence, local environmental conditions, and local distribution of
functional traits, among others. Specifically, although co-occurrence
can be driven by interactions (Cazelles et al., 2016), co-occurrence
alone is not a predictor of interactions (Blanchet et al., 2020; Thurman
et al., 2019), and therefore the lack of co-occurrence cannot be used to
infer the lack of a feasible interaction. Yet, recent results by Saravia
et al. (2021) strongly suggested that local (metaweb) realizations only
respond weakly to local conditions: instead, they reflect constraints
inherited by the structure of their metaweb. This sets up the core goal
of predictive network ecology as the prediction of metaweb structure, as
it is required to accurately produce downscaled, local predictions.

@@

Treating interactions as probabilistic (as opposed to binary) events is
a more nuanced and realistic way to represent them. Dallas et al. (2017)
suggested that most interactions (links) in ecological networks are
cryptic, \emph{i.e.} uncommon or hard to observe. This argument echoes
Jordano (2016): sampling ecological interactions is difficult because it
requires first the joint observation of two species, and then the
observation of their interaction. In addition, it is generally expected
that weak or rare interactions will be more prevalent in networks than
common or strong interactions (Csermely, 2004); this is notably the case
in food chains, wherein many weaker interactions are key to the
stability of a system (Neutel et al., 2002). In the light of these
observations, we expect to see an over-representation of low-probability
(hereafter rare) interactions under a model that accurately predicts
interaction probabilities.

Yet, the original metaweb definition, and indeed most past uses of
metawebs, was based on the presence/absence of interactions. Moving
towards \emph{probabilistic} metawebs, by representing interactions as
Bernoulli events (see \emph{e.g.} Poisot et al., 2016), offers the
opportunity to weigh these rare interactions appropriately. The inherent
plasticity of interactions is important to capture: there have been
documented instances of food webs undergoing rapid collapse/recovery
cycles over short periods of time (\emph{e.g.} Pedersen et al., 2017).
Furthermore, because the structure of the metaweb cannot be known in
advance, it is important to rely on predictive tools that do not assume
a specific network topology for link prediction (Gaucher et al., 2021),
but are able to work on generalizations of the network that capture
statistical processes giving it its structure. These considerations
emphasize why metaweb predictions should focus on quantitative
(preferentially probabilistic) predictions, and this should constrain
the suite of models that are appropriate for prediction. Binary
classifiers based on probabilities have an extremely robust methodology
to validate them, and this applies naturally to the prediction of
interactions (Poisot, 2023).

It is important to recall that a metaweb is intended as a catalogue of
all potential (feasible) interactions, which is then filtered for a
given application (Morales-Castilla et al., 2015). It is therefore
important to separate the interactions that happen ``almost surely''
(repeated observational data), ``almost never'' (repeated lack of
evidence \emph{or} evidence that the link is forbidden through
\emph{e.g.} trait mis-match), and interactions with a probability that
lays somewhere in between (Catchen et al., 2023). Although metawebs can
(and in practice likely do) include false positives, these are
statistically negligible compared to the false negatives. Furthermore,
Strydom et al. (2022) shows that t-SVD embedding is extremely robust to
(and able to detect) the presence of false positives. In a sense,
because most ecological interactions are elusive, we should consider the
direct consequences this has on sampling: once the common interactions
are documented, the effort required in documenting each rare interaction
will increase exponentially (Jordano, 2016). Recent proposals in other
fields relying on machine learning approaches emphasize the idea that
algorithms meant to predict, through the assumption that they
approximate the process generating the data, can also act as data
generators (Hoffmann et al., 2019). High quality observational data can
be used to infer core rules underpinning network structure, and be
supplemented with synthetic data coming from predictive models trained
on them, thereby increasing the volume of information available for
analysis. Indeed, Strydom, Catchen, et al. (2021) suggested that knowing
the metaweb may render the prediction of local networks easier, because
it fixes an ``upper bound'' on which interactions can exist. In this
context, a probabilistic metaweb represents an aggregation of
informative priors on the biological feasibility of interactions, which
is usually hard to obtain yet has possibly the most potential to boost
our predictive ability of local networks (Bartomeus, 2013; Bartomeus et
al., 2016). This would represent a departure from simple rules expressed
at the network scale (\emph{e.g.} Williams \& Martinez, 2000) to a view
of network prediction based on learning the rules that underpin
interactions \emph{and} their variability (Gupta et al., 2022).

@@

Graph (or network) embedding (Figure~\ref{fig-embedding}) is a family of
machine learning techniques, whose main task is to learn a mapping
function from a discrete graph to a continuous domain (Arsov \& Mirceva,
2019; Chami et al., 2022). Their main goal is to learn a low dimensional
vector representation of the graph (embeddings), such that its key
properties (\emph{e.g.} local or global structures) are retained in the
embedding space (Yan et al., 2005). The embedding space may, but will
not necessarily, have lower dimensionality than the graph. Ecological
networks are promising candidates for the routine application of
embeddings, as they tend to possess a shared structural backbone (see
\emph{e.g.} Bramon Mora et al., 2018), which hints at structural
invariants in empirical data. Assuming that these structural invariants
are common enough, they would dominate the structure of networks, and
therefore be adequately captured by the first (lower) dimensions of an
embedding, without the need to measure derived aspects of their
structure (\emph{e.g.} motifs, paths, modularity, \ldots).

@@

Rather than directly predicting biological rules (see \emph{e.g.}
Pichler et al., 2020 for an overview), which may be confounded by the
sparse nature of graph data, learning embeddings works in the
low-dimensional space that maximizes information about the network
structure. This approach is further justified by the observation, for
example, that the macro-evolutionary history of a network is adequately
represented by some graph embeddings {[}Random dot product graphs
(RDPG); see Dalla Riva \& Stouffer (2016){]}. In a recent publication,
Strydom et al. (2022) have used an embedding (based on RDPG) to project
a metaweb of trophic interactions between European mammals, and
transferred this information to mammals of Canada, using the
phylogenetic distance between related clades to infer the values in the
latent subspace into which the European metaweb was projected. By
performing the RDPG step on re-constructed values, this approach yields
a probabilistic trophic metaweb for mammals of Canada based on knowledge
of European species, despite a limited (\(\approx\) 5\%) taxonomic
overlap, and illustrates how the values derived from an embedding can be
used for prediction without being ``traits'' of the species they
represent.

@@

Graph embeddings \emph{can} serve as a dimensionality reduction method.
For example, RDPG (Strydom et al., 2022) and t-SVD {[}truncated Singular
Value Decomposition; Poisot et al. (2021){]} typically embed networks
using fewer dimensions than the original network {[}the original network
has as many dimensions as species, and as many informative dimensions as
trophically unique species; Strydom, Dalla Riva, et al. (2021){]}.
However, this is not necessarily the case -- indeed, one may perform a
PCA (a special case of SVD) to project the raw data into a subspace that
improves the efficacy of t-SNE {[}t-distributed stochastic neighbor
embedding; Maaten (2009){]}. There are many dimensionality reductions
(Anowar et al., 2021) that can be applied to an embedded network should
the need for dimensionality reduction (for example for data
visualization) arise. In brief, many graph embeddings \emph{can} serve
as dimensionality reduction steps, but not all do, neither do all
dimensionality reduction methods provide adequate graph embedding
capacities. In the next section (and Figure~\ref{fig-embedding}), we
show how the amount of dimensionality reduction can affect the quality
of the embedding.
